---
title: "Principal Components Analysis"
author: "Steve"
date: "October 30, 2014"
output: html_document
---

This video is in support of the following YouTube video:
Using prcomp and varimax for PCA in R www.youtube.com/watch?v=PSuvMBtvJcA 

******
Let's read in the file containing information on some wines. We want to analyze
their characteristics and see if there are ways we can group them. Wine afficionados
are concerned about these things.

```{r}
library(lattice)
my.wines <- read.csv("http://steviep42.bitbucket.org/YOUTUBE.DIR/wines.csv", header=TRUE)
```

Let's check out their correlations

```{r}
library(gclus)
my.abs     <- abs(cor(my.wines[,-1]))
my.colors  <- dmat.color(my.abs)
my.ordered <- order.single(cor(my.wines[,-1]))
cpairs(my.wines, my.ordered, panel.colors=my.colors, gap=0.5)
```

Next we will call a function to do the principal components. In this case we use prcomp
though some like to use princomp.

```{r}
my.prc <- prcomp(my.wines[,-1], center=TRUE, scale=TRUE)
screeplot(my.prc, main="Scree Plot", xlab="Components")
screeplot(my.prc, main="Scree Plot", type="line" )
```

Let's get a dotplot of the sorted loadings for the first component

```{r}
load    <- my.prc$rotation
sorted.loadings <- load[order(load[, 1]), 1]
myTitle <- "Loadings Plot for PC1" 
myXlab  <- "Variable Loadings"
dotplot(sorted.loadings, main=myTitle, xlab=myXlab, cex=1.5, col="red")
```

Let's get a dotplot of the sorted loadings for the second component

```{r}
sorted.loadings <- load[order(load[, 2]), 2]
myTitle <- "Loadings Plot for PC2"
myXlab  <- "Variable Loadings"
dotplot(sorted.loadings, main=myTitle, xlab=myXlab, cex=1.5, col="red")
```

Let's plot the Biplot and apply the varimax rotation

```{r echo=FALSE}
biplot(my.prc, cex=c(1, 0.7))
# Apply the Varimax Rotation
my.var <- varimax(my.prc$rotation)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
