---
title: "Express R - separate Steve Pittard"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyr)
library(ggplot2)
library(dplyr)
```

Consider the following simulated data. While it is a *toy* data set it is nonetheless demonstrative of some data I've seen previously in terms of format wherein some columns might encode two or more values. Check this out

```{r}
# Some example data
df <- data.frame(id=c("001","002","003","004","005","006"),
                 name=c("Tom Smith","Dee Ford","Mary Russell",
                                     "Frank Jones","Lisa Miller","Don Draper"),
                 age=c(32,43,22,39,50,60),
                 location=c("San Diego,CA","Los Angeles,CA","Monterey,CA",
                             "Portland,OR","Eugene,OR","Salem,OR"),
                 blood_pressure=c("120/90","140/85","135/110","125/102","128/92","145/118"),
                 stringsAsFactors = FALSE)

df
```


Note that the *name* column has first and last names and the *location* column encodes a city and state. And we also have a column with blood pressure information. We might want to split or *separate* these columns out into separate columns to facilitate summaries across states. Why would we to do this ? Well what if we wanted to aggregate across states or maybe create a bar chart of average age across states ? Or look at a systolic vs diastolic blood pressure across the States ?  As is, there is no convenient way to access the *State* or blood pressure info. The *separate* function will help. 


```{r}
(split.location <- separate(df,col=location,into=c("city","state"),sep=","))
```


```{r}
split.location %>% group_by(state) %>% summarize(ave=mean(age))
```
And we could create a bar chart

```{r}
split.location %>% group_by(state) %>% 
  summarize(ave=mean(age)) %>% 
  ggplot(aes(x=state,y=ave)) + geom_bar(stat="identity") + 
  ggtitle("Average Age Across State") + ylab("Average Age") + xlab("State")
```


Let's break out the blood pressure by systolic and disastolic:

```{r}
(split.location.bp <- split.location %>% 
  separate(col=blood_pressure,into=c("systolic","diastolic"),sep="\\/"))
```

Plot the systolic vs diastloic for each state. Use facets for this. Given that we don't have much data, the plot isn't
mind blowing in any way though we are just trying to illustrate how we reshape the data to suit our interests. 


```{r}
split.location.bp %>% ggplot(aes(x=diastolic,systolic)) + geom_point() + facet_wrap(~state)
```



